\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography \\ or \\ Charting antigenic drift and evolutionary success of seasonal influenza lineages

### 2013-01-07 flux.tex
Clarify virus effects.

%%% TITLE %%%
\title{\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography}

% Antigenic drift determines evolutionary success of seasonal influenza lineages

@@

Here, we instead assume a probabilistic interpretation in which an observed titer is normally distributed around its cartographic expectation with variance $\mdssd^2$,
\begin{equation} \label{hij}
	H_{ij} \sim \mbox{Normal}( \se_j - \delta_{ij}, \mdssd^2 ).
\end{equation}
Consequently, the likelihood of observing an exact titer given the placement of antigenic locations is 
\begin{equation} 
	\point(H_{ij}) = \phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{ \mdssd } \right),
\end{equation}
where $\phi(\cdot)$ represents the standard normal probability density function (PDF).
Previous BMDS has employed a sampling density truncated to strictly positive quantities since $d_{ij}$ are directly observed, non-negative quantities.  
In the antigenic setting, these remain random and can be negative \msc{I still don't really believe this} since neither $\se_j$ is known nor is $H_{ij}$ observed with much precision.

HI assays sometimes show no inhibition at all measured titrations, e.g.\ a measurement can be reported as `$<$40'.
In this case, the likelihood of observing the threshold measurement follows the cumulative density of the lower tail of the normal distribution
\begin{equation} 
	\threshold(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{ \mdssd } \right),
\end{equation}
where $\Phi(\cdot)$ represents the standard normal cumulative distribution fucntion (CDF).
Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:40 dilution and no inhibition at 1:80 dilution.
Rather than taking the HI titer as 40, we can instead treat this as an interval measurement, assuming that the exact titer for inhibition would occur somewhere between 40 and 80.
HI titers are usually reported as the highest titer that successfully inhibits virus binding, so that in this case, we calculate the likelihood of an interval measurement as
\begin{equation} 
	\interval(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j + 1 }{ \mdssd } \right) - \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{\mdssd} \right).
\end{equation}
These likelihoods are illustrated in Figure~\ref{hij_likelihood}.
Throughout our analyses, we use interval likelihoods $\interval$ rather than point likelihoods $\point$ unless otherwise noted.

@@

We calculate the overall likelihood by multiplying probabilities of individual measurements
\begin{equation} 
	L(\viruses,\sera) = \prod_{(i,j) \in \cal I} f(H_{ij}),
\end{equation}
using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$, and we start by assuming independent, diffuse normal prior on virus and serum locations
\begin{eqnarray}
	\virus_i &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}) \nonumber \\
	\serum_j &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}),
\end{eqnarray}
with $\boldsymbol{\mu} = (0,\ldots,0)^{\prime} $ and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $10000$.

@@

The preceding model represents immunological distance as a drop in titer against the most reactive comparison for a particular serum.
However, this model may be biased in some circumstances.
In one example, if a particular serum $j$ is only measured against distant viruses, its maximum titer will be artificially low and the likelihoods concerning this serum will appear poor. 
To address this issue, we relax the assumption of fixed $\se_j$ values and treat the expected log$_2$ titer when $\delta_{ij}=0$ as a random variable.
In this case, $H_{ij}$ still follows equation \ref{hij} with expectation $\se_j - \delta_{ij}$, but the vector of `serum effects' $\ses = (\se_1,\ldots,\se_{\sn})$ is random and estimated rather than fixed.
We assume that $\se_j$ values are hierarchically distributed according to a normal distribution.   
We take an Empirical Bayesian approach is specifying the mean and variance of this distribution, set to the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
This formulation assumes that particular sera are more reactive in general than other sera.

Additionally, we follow the same logic and assume that some virus isolates are more reactive in general than other virus isolates.
With virus reactivity included, the expectation of $H_{ij}$ becomes $(\ve_i+\se_j)/2 - \delta_{ij}$ and the vector of `virus effects' $\ve_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\ves$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.

With these effects included, Metropolis-Hastings proposals additionally incorporate transition kernels for individual serum effects $\se_j$ and virus effects $\ve_i$. \msc{What are $\ve_i$?  These have not been defined nor used.}

@@

As presented, virus and serum locations $\viruses$ and $\sera$ accurately characterize antigenic distances between virus strains based on titers from pairwise HI assays.
However, these distances represent only a local description of antigenic space and do not provide a global picture.
An example of this phenomena is shown in Figure~\ref{schematic_map}.
In this case it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar or antigenically divergent. 
This presents an issue of model identifiability; multiple antigenic maps give the same likelihood of observing the serological data.
In order to achieve more interpretable antigenic locations we impose a weak prior on global locations.
In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
To capture this, we replace the diffuse normal prior on virus locations with a prior where virus locations follow a normal distribution in which the expectation of dimension 1 increases with date of virus sampling
\begin{equation}
	\virus_i \sim \mathrm{Normal}(\mu, \Sigma),
\end{equation}
where $\mu = \twomatrix{\beta \, t_i}{0}$ and $\Sigma = \fourmatrix{\driftsd^2}{0}{0}{\driftsd^2}$ for $P=2$, and $t_i$ is the difference in time between virus $i$ and the earliest sampled virus.
Thus, this model assumes that the virus population drifts in a line across the antigenic map at rate $\beta$.
The parameter $\driftsd$ determines the breadth of the virus population cloud at each point in time.
We assume that both the prior slope $\beta$ and the prior precision $1/\driftsd^2$ follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$.
We further introduce Metropolis-Hastings proposals that scale $\beta$ and $\driftsd$.

>>>>

%%% TITLE %%%
\title{\vspace{1.0cm} \LARGE \bf 
Revealing the competitive dynamics of influenza viruses through evolutionary cartography \\
or \\
Charting antigenic drift and evolutionary success of seasonal influenza lineages
}

@@

Here, we instead assume a probabilistic interpretation in which an observed titer is normally distributed around its cartographic expectation with variance $\mdssd^2$,
\begin{equation} \label{hij}
	H_{ij} \sim \mbox{Normal}( \se_j - \delta_{ij}, \mdssd^2 ).
\end{equation}
Consequently, the likelihood of observing an exact titer given the placement of antigenic locations is 
\begin{equation} 
	\point(H_{ij}) = \phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{ \mdssd } \right),
\end{equation}
where $\phi(\cdot)$ represents the standard normal probability density function (PDF).
Previous BMDS has employed a sampling density truncated to strictly positive quantities since $d_{ij}$ are directly observed, non-negative quantities.  
In the antigenic setting, these remain random and can be negative since neither $\se_j$ is known nor is $H_{ij}$ observed with much precision.

HI assays sometimes show no inhibition at all measured titrations, e.g.\ a measurement can be reported as `$<$40'.
In this case, the likelihood of observing the threshold measurement follows the cumulative density of the lower tail of the normal distribution
\begin{equation} 
	\threshold(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{ \mdssd } \right),
\end{equation}
where $\Phi(\cdot)$ represents the standard normal cumulative distribution function (CDF).
Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:40 dilution and no inhibition at 1:80 dilution.
Rather than taking the HI titer as 40, we can instead treat this as an interval measurement, assuming that the exact titer for inhibition would occur somewhere between 40 and 80.
HI titers are usually reported as the highest titer that successfully inhibits virus binding, so that in this case, we calculate the likelihood of an interval measurement as
\begin{equation} 
	\interval(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j + 1 }{ \mdssd } \right) - \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{\mdssd} \right).
\end{equation}
These likelihoods are illustrated in Figure~\ref{hij_likelihood}.
Throughout our analyses, we use interval likelihoods $\interval$ rather than point likelihoods $\point$ unless otherwise noted.

@@

We calculate the overall likelihood by multiplying probabilities of individual measurements
\begin{equation} 
	L(\viruses,\sera) = \prod_{(i,j) \in \cal I} f(H_{ij}),
\end{equation}
using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$, and we start by assuming independent, diffuse normal prior on virus and serum locations
\begin{eqnarray}
	\virus_i &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}) \nonumber \\
	\serum_j &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}),
\end{eqnarray}
with $\boldsymbol{\mu} = (0,\ldots,0)^{\prime}$ and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $10000$.

@@

The preceding model represents immunological distance as a drop in titer against the most reactive comparison for a particular serum.
However, this model may be biased in some circumstances.
In one example, if a particular serum $j$ is only measured against distant viruses, its maximum titer will be artificially low and the likelihoods concerning this serum will appear poor. 
To address this issue, we relax the assumption of fixed $\se_j$ values and treat the expected log$_2$ titer when $\delta_{ij}=0$ as a random variable.
In this case, $H_{ij}$ still follows equation \ref{hij} with expectation $\se_j - \delta_{ij}$, but the vector of `serum effects' $\ses = (\se_1,\ldots,\se_{\sn})$ is random and estimated rather than fixed.
We assume that $\se_j$ values are hierarchically distributed according to a normal distribution.   
We take an Empirical Bayesian approach in specifying the mean and variance of this distribution, set to the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
This formulation assumes that particular sera are more reactive in general than other sera.

Additionally, we follow the same logic and assume that some virus isolates are more reactive than other virus isolates and include a `virus effect' $\ve_i$ representing the general level of reactivity across HI assays. 
With virus reactivity included, the expectation of $H_{ij}$ becomes $(\ve_i+\se_j)/2 - \delta_{ij}$ and the vector of virus effects $\ve_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\ves$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.

With these effects included, Metropolis-Hastings proposals additionally incorporate transition kernels for individual serum effects $\se_j$ and virus effects $\ve_i$.

@@

As presented, virus and serum locations $\viruses$ and $\sera$ accurately characterize antigenic distances between virus strains based on titers from pairwise HI assays.
However, these distances represent only a local description of antigenic space and do not provide a global picture.
An example of this phenomena is shown in Figure~\ref{schematic_map}.
In this case it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar or antigenically divergent. 
This presents an issue of model identifiability; multiple antigenic maps give the same likelihood of observing the serological data.
In order to achieve more interpretable antigenic locations we impose a weak prior on global locations.
In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
To capture this, we replace the diffuse normal prior on virus locations with a prior where virus locations follow a normal distribution in which the expectation of dimension 1 increases with date of virus sampling
\begin{equation}
	\virus_i \sim \mathrm{Normal}(\boldsymbol{\mu}, \boldsymbol{\Sigma}),
\end{equation}
where $\boldsymbol{\mu} = (\beta \, t_i, 0)^{\prime}$ and $\boldsymbol{\Sigma} = \fourmatrix{\driftsd^2}{0}{0}{\driftsd^2}$ for $P=2$, and $t_i$ is the difference in time between virus $i$ and the earliest sampled virus.
Thus, this model assumes that the virus population drifts in a line across the antigenic map at rate $\beta$.
The parameter $\driftsd$ determines the breadth of the virus population cloud at each point in time.
We assume that both the prior slope $\beta$ and the prior precision $1/\driftsd^2$ follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$.
We further introduce Metropolis-Hastings proposals that scale $\beta$ and $\driftsd$.
